---
title: "Cyclistic Bikes"
author: "Jhon Montesdeoca"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---
## Google Data Analytics Capstone Project: How do annual members and casual riders use Cyclistic bikes differently?

## Phase 1: Ask
Note: This phase is where we identify, define and ask the right question to solve our data problem.

Cyclistic launched in 2016 and has had much success offering their bike-share services. The program has expanded to have close to 6,000 bicycles that are also geo-tracked and locked into a network of over 600 stations across Chicago. The bikes can be unlocked at one station and return to another without out any issues. Cyclistic was able to adapt to a changing market and keep things fresh by offering different types of riding types like: reclicing bikes, hand tricycles, and cargo bikes. Cyclistic users are more likely to ride for leisure, but about 30% use them to commute to work each day.                                                                                                                          The director of the marketing team Lily Moreno belives that converting casual riders into members will help the company grow. However, before making any further decisions the marketing team needs to understand what makes the differences are between the casual riders and member riders, why would a casual rider convert to a membership, and how digital media could affect their marketing approach. 

#### Business Task

Analyze the most recent 12 month Cyclistic Customer Data in order to understand how annual members and casual riders use Cyclistic bikes differently, and provide a data-driven strategy that will turn casual riders into annual members.

#### Business Objectives
* Organize the data to clearly spot trends about annual and casual riders.
* Run a comprehensive analysis for finding the most optimized marketing strategy.
* Create supporting visualizations to communicate the data.

#### Deliverables
* A clear statement of the business task
* A description of all data sources used
* Documentation of any cleaning or manipulation of data
* A summary of your analysis
* Supporting visualizations and key findings
* Your top three recommendations based on your analysis

#### Stakeholders
* Lily Moreno: The director of marketing and your manager.
* Cyclistic executive team: The notoriously detail-oriented executive team will decide whether to approve the recommended marketing program.
* Cyclistic marketing analytics team: A team of data analysts who are responsible for collecting, analyzing, and reporting data that helps guide Cyclistic marketing strategy.

## Phase 2: Prepare
Note: In the prepare phase, we analyze the quality(ROCC) of the data and the limitations it has.

* Data is contained in a public dataset which can be found at: https://divvy-tripdata.s3.amazonaws.com/index.html. It contains 35 ZIP files.
* Data is provided by Motivate International Inc. under this license agreement.
* Data was collected by the customers activity on Cyclistic but customers were kept anonymous they all were issued a unique ID.
* I used the 2019 Q2,Q3,Q4 and the 2020 Q1 since they are the most recent 12 month data available on the data set.

#### Limitations of Dataset
* Data is collected 2 years ago, so it is not the most up to date data.
* The data is provided by a third-party so it may not be as reliable as first-party data.
* Although all of the quarters have large sample sizes some quarters differ vastly from each other in terms of sample size.

#### ROCC Check
* Reliable: High, the data has a very large sample size( about a million rows on the trip_id column for each quarter.)
* Original: Low, the data is provided by a third-party called Motivate International Inc.
* Comprehensive: High, the data has more traits than needed for this analysis.
* Current: Med, the data is 2 years old it is not the most up to date data but it isn’t outdated either.
* Cited: Low, the data is collected from a third-party.

#### Data Selection

Divvy_Trips_2019_Q2.csv                                                                                                 
Divvy_Trips_2019_Q3.csv                                                                                                 
Divvy_Trips_2019_Q4.csv                                                                                                 
Divvy_Trips_2020_Q1.csv                                                                                                 

These files where inported into R studio and combined into a single CSV file:
all_trips_v2

#### Tools

Process: R Studio was used to clean, filter, sort and transform the data.
Analyze: R Studio was used for analysis.
Visualization: Ggplot2 package in R was used and Tableau for further visuals

## Phase 3: Process
This is the phase where the data is cleaned and transformed to be ready for analysis.

#### Setting up my environment
Notes: setting up my R environment by loading the 'tidyverse', 'lubridate', and 'ggplot2' packages
```{r loading packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

#### Loading files, cleaning, sorting, and filtering data
```{r loading csv files to condense into one file to clean, message=FALSE, warning=FALSE}
# Upload Divvy datasets (csv files)
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

```{r}
# Compare column names each of the files
# While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2020)
```

```{r}
# Rename columns  to make them consistent with q1_2020 (as this will be the supposed going-forward table design for Divvy)

(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))

# Inspect the data frames and look for incongruities
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)

# Convert ride_id and rideable_type to character so that they can stack correctly
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

# Stack individual quarter's data frames into one big data frame
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)

# Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))

# Inspect the new table that has been created
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.
tail(all_trips)  #See last 6 rows of data frame.
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numeric

# Check to make sure the proper number of observations were reassigned
table(all_trips$member_casual)

# Reassign to the desired values (we will go with the current 2020 labels)
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))


# Add columns that list the date, month, day, and year of each ride
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

# Add a "ride_length" calculation to all_trips (in seconds)
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

# Inspect the structure of the columns
str(all_trips)

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
is.factor(all_trips$ride_length)

all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))

is.numeric(all_trips$ride_length)

# The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by              Divvy or ride_length was negative
# We will create a new version of the dataframe (v2) since data is being removed
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

## Phase 4: Analysis
Calculated the mean, median, maximum value and minimum value of the cleaned data frame.
```{r}
# Descriptive analysis on ride_length (all figures in seconds)
# One line using summary() on the specific attribute
summary(all_trips_v2$ride_length)
```

#### Data Aggregation
Compared the members and casual users.
```{r}
# Compare members and casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

# See the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
#### Analysis
Fixed the days of the week into order, ran the average ride time by each day for members vs casual users, analyzed the ridership data by type and weekday.
```{r}
# Notice that the days of the week are out of order. Let's fix that.
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Now, let's run the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

#analyze ridership data by type and weekday
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```

## Phase 5: Share
Creating the visuals using ggplot2.

#### Number of rides
Here, we see the comparison of number of rides by day for member vs casual riders

```{r visualize number of trips by rider type, message=FALSE, warning=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(y = "Number of Rides",x = "Weekday", fill = "Member Casual")
  
```


#### Average trip duration
Here, we see the average duration of rides by day for member vs casual riders

```{r visualize ride duration by rider type, message=FALSE, warning=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(y = "Average Duration(seconds)",x = "Weekday", fill = "Member Casual")
```

#### Exporting The Data
Exported data as CSV file
```{r}
#create DF and export CSV for Visuals
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = '/Users/jhonni/R/Cyclistic_Proj/avg_ride_length.csv')
```

## Phase 6: Act
This is the phase where we carefully go through our data problem and the analysis we made to make a data-driven decision.

Based on the data and comparisons on the charts we can see that members use Cyclistic more throughout the week and the ride count is hgiher than a casual user, probably used to commute to work since its Monday - Friday.

However, casual users are noticeably using the bikes for longer periods of time than members which is another indication that member users use the bikes for commuting purposes. Although the bike ride count is higher, the duration is longer with casual riders.

## Conclusion

My recommendation to the stake holders would be to restructure a marketing campaign where we prioritize ride duration over ride count. If the goal is to convert casual riders to members, having a benefit where they can use the bikes for longer as a member would be beneficial to the casual rider that is shows to use it longer when it's for fun. The amount of casual riders converting to members is sure to increase.

##                                                      Thank You!
